---
bibliography: [references.yaml, references.bib]
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| label: setup
#| include: false
source("packages.R")
library(terra)
library(sf)
library(ggplot2)
library(patchwork)
```

# Materials and Methods

## Sampling Design

The FLEA (Flanders Ecosystem Accounting) maps are raster-based maps of Flanders that are produced with a three year interval.
Each raster cell measures 10 m by 10 m.
The goal for the sampling design is to provide both validation of ecosystem changes between map datums and unbiased estimates of the changes in area.

As a first step, we construct the sampling frame by combining the $p$ maps into maps of change following the procedure in @gong2023.
Maps were available for $p = 3$ years: 2016, 2019 and 2022.
For each ecosystem class, we considered the union of the three periods (any raster cell where the ecosystem was present in at least one of the $K$ periods) and encoded each raster cell into "stable" (`111`), "gained" (`001`, `011`), "lost" (`110`, `100`), and "other complex patterns" (`101`, `010`), yielding a maximum of 4 strata per ecosystem type.
Each of the $p$ maps has categorical raster values encoding for $K = 16$ ecosystem types.
Thus, we obtain $16 \times 4 = 64$ strata.

Within each stratum, we select a spatially balanced sample of raster cells for which we make use of the same `GRTS` master sample that is used for the carbon monitoring network (`Cmon`) [@sleutel2020] and which is redistributed as a raster map [@vancalster2024].
The coordinate reference system, resolution and origin of the raster map are the same as used for the ecosystem maps, which makes them compatible and easy in use for selection of probabilistic spatially balanced samples.
Each raster cell of the `GRTS` master sampling raster map holds a ranking number.
These numbers can be used to draw spatially balanced samples of size $n$ for strata within the bounding box by sorting the ranks from low to high and selecting the $n$ lowest ranks (after cropping the raster file with the boundaries of the stratum).

Because spatially balanced samples tend to be well-spread across the spatial domain (avoiding clustered selection of sampling locations), we expect that within a stratum, say "forest gained", the selected sampling units will naturally reflect the area shares of constituent ecosystem transitions (see @tbl-conversions).
For instance, if half of the area "forest gained" is a transition from grassland to forest, we expect half of the sampling units to be located within that specific transition.
It is however not guaranteed that every transition will be part of the sample due to finite sample sizes and relative small area shares of some transitions.

As a rule, we allocated 80 sampling units to the "stable" class for each ecosystem and 240 sampling units to the classes representing changes in ecosystem type (i.e. "gained", "lost", "other complex pattern").
Only change classes that represent at least 10 ha are considered.
The allocation of the 240 sampling units to the "gained", "lost" and "other complex pattern" substrata is done proportional to the map area of these substrata, with the constrained that each substratum receives at least 40 sampling units.

```{r}
#| label: tbl-conversions
#| include: true
#| tbl-cap: Allocation of sample size among strata. The conversions within each change class (gain, loss or other complex pattern) are also given. The numbers between brackets indicate the breakdown of the total sample size within that stratum for each specific conversion.
withr::with_dir(
  rprojroot::find_root(rprojroot::is_git_root),
  {
    library(targets)
    Sys.setenv(TAR_PROJECT = "validation_sample")
    tm <- tar_read(temporal_map)
    vs <- tar_read(validation_sample)
    vp50 <- tar_read(validation_polygons_50)
    tar_load(catstable)
  }
)
vs_conversions <- terra::extract(x = tm, y = vect(vs), raw = FALSE, bind = TRUE)
vs_conversions_df <- values(vs_conversions) |>
  as_tibble() |>
  rename(conversion = label) |>
  mutate(
    stratum_code = stringr::str_extract(stratum_name, "^bc_(\\d+)_changecat", 1)
  ) |>
  left_join(
    catstable |>
      select(value, ecosystem = label) |>
      mutate(value = as.character(value)),
    by = join_by(stratum_code == value)
  ) |>
  count(stratum_code, ecosystem, changecat, conversion, name = "n_central_cell")

vs_conversions_df |>
  mutate(stratum = paste(stratum_code, changecat)) |>
  group_by(stratum, ecosystem) |>
  arrange(desc(n_central_cell)) |>
  summarize(
    n_total = sum(n_central_cell),
    conversions = xfun::join_words(
      sprintf("%s (%s)", conversion, n_central_cell)
    ),
    .groups = "drop"
  ) |>
  select(ecosystem, stratum, n_total, conversions) |>
  kable(
    booktabs = TRUE
  ) |>
  kableExtra::collapse_rows(valign = "top")
```


## Response Design

### Definition of Ecosystems

Level 1 of the ecosystem typology of EU Regulation 2024/3024 is a standardized and mandatory format for EU member states' ecosystem accounts.
For levels 2 and 3 Eurostat developed a proposal based on the Corine- and EUNIS-typologies.
Member states are encouraged to adopt these extra levels, but can also choose to adapt them according to their own national policy priorities or data availability.
@tbl-typology presents an ecosystem typology - version 1.3, March 2026 - for the Flemish ecosystem extent account.

Level 1 contains 12 ecosystems, level 2 35 subsystems.
For the Flemish accounts, level 2 of ecosystem '1. Settlements and other artificial areas' is based on land cover classes: buildings, unbuilt sealed soil, high green, low green, standing water and uncertain land cover (see tbl-settlement).
The latter is mostly a mixture of low green and limited (< 76%) sealed soil.
The ecosystem extent validation includes validation of the settlement level 2 subdivisions.

At level 3 of the typology these land covers are detailed further according to various land use sectors (residential, commercial & industrial, transport, recreational, ...).
Level 3 is not used for ecosystem extent mapping and validation.
For the other ecosystems the EU level 2-typology is mostly followed, to the extent that data are available. 

```{r}
#| label: tbl-typology
#| results: asis
#| tbl-cap: EU (Maes et al., 2013; EU ecosystem typology, Eurostat 2024) and Flemish level-1 ecosystem type definitions. Flemish column is in Dutch, pending review and translation.

ecosystem_types <- data.frame(
  `Ecosystem Type level 1` = c(
    "1. Settlements and other artificial areas",
    "2. Cropland",
    "3. Grassland (pastures, semi-natural and natural grasslands)",
    "4. Forest and woodland",
    "5. Heathland and shrub",
    "6. Sparsely vegetated ecosystems",
    "7. Inland wetlands",
    "8. Rivers and canals",
    "9. Lakes and reservoirs",
    "10. Marine inlets and transitional waters",
    "11. Coastal beaches, dunes and wetlands",
    "12. Marine ecosystems (coastal waters, shelf and open ocean)"
  ),
  `EU description and source` = c(
    r"{**Description:** This class represent human habitats, i.e. ecosystems that are strongly modified by people and that are characterised by buildings and other man-made structures. It includes dispersed and dense residential, industrial, commercial, and transport areas, urban green areas, mineral and gravel extraction sites, dumping and construction sites. The class excludes peat extraction sites (classified under 'Inland wetlands'). The class has also been labelled 'urban ecosystems'. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Croplands are food production areas including both intensively managed cropland, extensively managed cropland, and multifunctional areas. They include areas with perennial and annual crops, and agro-ecosystems with significant coverage of natural vegetation (agricultural mosaics). Cropland may include agro-forestry areas such as cork and holm oak stands (see respective classes at level 2). Small semi-natural elements which are associated with agricultural land use, such as hedges, ponds, grassy margins etc., are considered part of this ecosystem type. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Grassland covers areas dominated by herbaceous vegetation (including in particular grasses and forbs, but also mosses and lichens) of two kinds -- modified grasslands (sown and used for grazing or hay and silage production) and (semi-)natural (extensively managed) grasslands. Grassland also includes some agroforestry area used for grazing, such as cork and holm oak stands or wooded pasture. Small semi-natural elements which are associated with agricultural land use, such as hedges, ponds, grassy margins etc. are considered part of this ecosystem type. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Forests and woodlands include tree-dominated ecosystems with a canopy cover generally over 30% (in the temperate climate zone) or 10% (in boreal and Mediterranean climate zones) or vegetation where present trees can reach these thresholds in situ. For boreal and Mediterranean forest types a lower threshold is advisable due to extreme climatic conditions, which naturally limit the growth of trees and relative canopy cover. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Heathland and shrub are areas with vegetation dominated by shrubs or dwarf shrubs, which may include dispersed trees with a canopy cover below 30%. They include natural ecosystems of predominantly harsh conditions as well as secondary ecosystems created by extensive human use. The class includes heathlands and sclerophyllous vegetation. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Sparsely or unvegetated ecosystems have a low density of vegetation, with typically at least 70% of barren soil. They include degraded sparsely or degraded non-vegetated areas as well as ecosystems under extreme natural conditions which may be traditionally grazed. They include bare rocks, glaciers, inland dunes and sand plains. In spite of low vegetation density, the ecosystems may contain high biodiversity of species adapted to an extreme environment. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Inland wetlands are areas that are year-round or seasonally strongly affected by water, in the form of temporary flooding or groundwater levels close to surface. This class includes natural, semi-natural or modified inland marshes as well as mires, bogs and fens, but excludes seasonally flooded grasslands and heathlands, for example. It also includes peat extraction sites. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Permanent freshwater inland surface waters of linear character. These include natural water courses, such as rivers, streams etc., as well as anthropogenic structures built for transportation, drainage or water supply purposes, i.e., canals, ditches etc. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Permanent freshwater inland surface waters composed of primarily non-linear water bodies. These include natural water bodies (lakes) as well as anthropogenic water bodies developed mainly for water supply or energy generation purposes. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** Marine inlets and transitional waters are ecosystems on the land-water interface under the influence of tides and with salinity higher than 0.5 permille. They include lagoons, estuaries and other transitional waters. **Source:** Based on Maes et al., 2013.}",
    r"{**Description:** The class coastal areas refers to land-based ecosystems close to the sea, with marine influences, such as salt spray and saline groundwater, and that may be flooded during high tide or extreme events. The class includes beaches, dunes (which may include wetland areas below the MMU in between rows of dunes) and coastal saltmarshes and salines. **Source:** Completes terrestrial ecosystem types from Maes et al., 2013.}",
    r"{**Description:** Marine ecosystems include all marine areas in the sea extent at low tide level or below mean sea level. This ranges from near-shore ecosystems to deep water marine ecosystems. Aligned with accounting principles, the ecosystem includes the whole water column including the seabed and the pelagic zone. **Source:** Completes Maes et al. 2013 with a marine ecosystem type.}"
  ),
  `Flemish description and source (in Dutch, to be reviewed, translation after review)` = c(
    r"{**1. Woongebieden en andere kunstmatige gebieden**
**Beschrijving:** Alle gebieden met een in hoofdzaak artificiele bodembedekking ingevolge menselijke activiteiten: steden, dorpen, villawijken, industrieterreinen, recreatiegebieden, campings, (spoor)wegen, vliegvelden, serres, parkings, actieve groeves en stortplaatsen. Het bebouwd gebied omvat hierbij ook alle terreinen die aan deze gebouwen, wegen ... gelinkt zijn, zoals tuinen, parken, kerkhoven, volkstuintjes, sportvelden, terrils, groeves, waterbassins en recent vergraven terreinen (De Saeger et al., 2026). De definitie komt grotendeels overeen met het ruimtebeslag, 'ruimtebeslag', met uitzondering van waterlichamen (Poelmans et al., 2022). Stilstaande waters op percelen die tot het ruimtebeslag behoren (bv. tuinvijvers) of met een recreatieve functie (bv. recreatie) worden tot de kunstmatige gebieden gerekend. Stromend water (bv. rivieren, kanalen en grachten) maakt geen deel uit van het ruimtebeslag tenzij ze niet bovengronds waarneembaar zijn, bv. door inbuizing. Verharding wordt steeds bij ruimtebeslag gerekend.
**Bron:** Landgebruiksbestand Vlaanderen (Poelmans et al., 2022, para. 6.2) en de Jaarlijkse Bodemafdekkingskaart (JaarBAK) van het departement Omgeving.}",
    r"{**2. Akkerland**
**Beschrijving:** Alle eenjarige en meerjarige professionele landbouwteelten, inclusief tijdelijk grasland op percelen die de voorbije vijf jaar ook andere teelten dan grasland kenden. Dit omvat ook kweek van bomen en sierplanten, kleinfruit en laagstamboomgaarden. Teelten binnen een gebouw of permanente serre zijn deel van de landbouwinfrastructuur en behoren tot ecosysteem 1. Kleine landschapselementen op de akkerpercelen zoals hagen en houtkanten maken deel uit van het akkerland. Enkel landbouwpercelen geregistreerd in de inventaris van het departement Landbouw en Zeevisserij komen in aanmerking, niet de percelen beheer door hobbyboeren (Poelmans et al., 2022). Hoogstamboomgaarden buiten ruimtebeslag horen bij ecosysteem 3. Grasland. Moestuinen op residentiele of andere percelen binnen ruimtebeslag horen bij ecosysteem 1.
**Bron:** Landgebruiksbestand Vlaanderen, niveau 1, klassen akker, laagstamboomgaard en tijdelijk grasland (Poelmans et al., 2022) en Landbouwgebruikspercelen. Voor een specifieke lijst van akkerteelten, zie bijlage.}",
    r"{**3. Grasland (weiland, halfnatuurlijk en natuurlijk grasland)**
**Beschrijving:** Open biotopen gedomineerd door lage en hoge grassen en door lage, meerjarige kruiden. Vele graslanden zijn geperceleerd en meestal in professioneel of hobby-landbouwgebruik. Graslanden kennen meestal een jaarlijks beheer van maaien en/of begrazen. In verlaten graslanden of graslanden onder zeer extensieve begrazing kan een lage bedekking aan (jonge) bomen, struiken en bramen voorkomen (De Saeger et al., 2026). Hoogstamboomgaarden, binnendijkse zilte vegetaties, ingezaaide graslanden, door pitrus gedomineerde vegetaties met weinig moerasplanten, ... Tijdelijk grasland in professioneel landbouwgebruik dat deel uitmaakt van een teeltrotatie en dat niet de status 'BG - blijvend grasland' heeft, maakt deel uit van ecosysteem 2. Tot grasland behoren eveneens ruigten en pioniervegetaties: open biotopen gedomineerd door lage of hoge kruiden, soms kunnen ze een aandeel open bodem bevatten (De Saeger et al., 2026). Graslandvegetatie op percelen die tot het ruimtebeslag horen, zoals in tuinen, parken en bedrijventerreinen, behoren tot ecosysteem 1.
**Bron:** Landgebruiksbestand Vlaanderen niv. 1 (Ruigte, Halfnatuurlijk grasland, Boomgaard hoogstam, Blijvend grasland, Overig laag groen en Overig (indien niet verhard)); Landbouwgebruikspercelen met graslandteelt en status 'BG-Blijvend Grasland', Biologische Waarderingskaart, Groenkaart en Jaarlijkse Bodemafdekkingskaart. Voor een specifieke lijst van graslandvegetatie en graslandteelten, zie bijlage.}",
    r"{**4. Bossen en bosgebieden**
**Beschrijving:** Vegetatie gedomineerd (kroonprojectie >= 50%) door bomen; inclusief recente kapvlakten, jonge (her)aanplant, braamstruweel en hakhout, die niet op percelen van het ruimtebeslag (ecosysteem 1) liggen en geen professionele landbouwfunctie hebben. Aanplanten voor korte omloop hakhoutteelt (energiehoutakkers), kerstboomaanplantingen en kwekerij van (sier)bomen en struiken vallen onder ecosysteem 2 (De Saeger et al., 2026).
**Bron:** Landgebruiksbestand Vlaanderen niv. 1 (Struweel, Loofbos, Populieren, Naaldbos, Alluviaal bos en Overig hoog groen) (Poelmans et al., 2022). Wanneer bos of bomen voorkomen in gebieden die tot het ruimtebeslag behoren of in professioneel landbouwgebruik zijn (bv. parken, tuinen, houtkanten), horen ze bij ecosystemen 1, 2 of 3. Doordat de methode van het Landgebruiksbestand Vlaanderen wordt gevolgd, wordt bij een afwijking tussen de Boswijzer en de Biologische Waarderingskaart, de Boswijzer gevolgd. Voor een specifieke lijst van bosvegetatie, zie bijlage.}",
    r"{**5. Heide en struiken**
**Beschrijving:** Vegetatie gedomineerd door dwergstruiken (Calluna, Erica, Vaccinium, Genista) in een heidegebied, die niet op percelen van het ruimtebeslag (ecosysteem 1) ligt of op percelen met een professionele landbouwfunctie. Inclusief en plag- of brandplekken, en soortenarme, nagenoeg monospecifieke vegetaties gedomineerd door pijpenstrootje en/of bochtige smele, met afwezigheid of lage bedekking aan 'heideplanten' (struikhei, gewone dophei, beenbreek, veenpluis, ...). Meestal maken die deel uit van een groter 'heidegebied' (De Saeger et al., 2026).
**Bron:** Landgebruiksbestand Vlaanderen, niveau 1 (Heide) en Biologische Waarderingskaart. Voor een specifieke lijst van heidevegetatie, zie bijlage.}",
    r"{**6. Ecosystemen met schaarse begroeiing**
**Beschrijving:** Landduinen en zeer droge zandgronden met open stuifzand en/of (deels) begroeid met korstmossen, haarmossen, grijs kronkelsteeltje, buntgras, zandstruisgras, zandzegge, heidespurrie, vroege haver, zilverhaver, struikhei (De Saeger et al., 2026). Enkel indien zij niet op percelen van het ruimtebeslag (ecosysteem 1) liggen of op percelen met een professionele landbouwfunctie hebben.
**Bron:** Landgebruiksbestand Vlaanderen, niveau 1 (Heide) en Biologische Waarderingskaart. Voor een specifieke lijst van landduinvegetatie, zie bijlage.}",
    r"{**7. Wetlands in het binnenland**
**Beschrijving:** Vegetatie met > 30% moerasplanten (helofyten en hydrofyten); watertafel een groot deel van het jaar nabij of boven het maaiveld, in zoverre zij niet op percelen van het ruimtebeslag (ecosysteem 1) liggen of op percelen met een professionele landbouwfunctie. Veel voorkomende en indicatieve moerasplanten zijn: riet, liesgras, lisdodde, egelskop, mattenbies, heen, grote zeggen (moeras-, oever-, snavel-, scherpe zegge), holpijp, galigaan, waterdrieblad en wateraardbei (De Saeger et al., 2026).
**Bron:** Landgebruiksbestand Vlaanderen, niveau 1 (Moeras) en Biologische Waarderingskaart. Voor een specifieke lijst van moerasvegetatie, zie bijlage.}",
    r"{**8. Rivieren en kanalen**
**Beschrijving:** Natuurlijke waterlopen (rivieren en beken) en kunstmatige waterlopen (kanalen en grachten), die niet onder getijdeninvloed staan en waarvan de lengte in een richting minstens 10 m en in de dwarse richting 3m bedraagt (Poelmans et al., 2022, p. 10). Waterlopen die bovengronds niet waarneembaar zijn, bv. omdat ze zijn ingebuisd, worden tot ecosysteem 1 gerekend. Waterlopen onder getijdeninvloed worden tot ecosysteem 10 gerekend.
**Bron:** Landgebruiksbestand Vlaanderen, niveau 1 (Water). Deze laag werd aangepast voor de stilstaande waters (toegerekend aan ecosysteem 9, o.b.v. Watervlakkenkaart) en voor de getijdengevoelige waterlopen (toegerekend aan ecosysteem 11, o.b.v. Moneos-data).}",
    r"{**9. Meren en waterbekkens**
**Beschrijving:** Duidelijk begrensd stilstaand open water dat gedurende meerdere jaren een groot deel van het vegetatieseizoen aanwezig is en zich niet op percelen met ruimtebeslag (ecosysteem 1) bevindt. Het waterlichaam wordt als 'stilstaand' beschouwd als er geen visueel waarneembare stroming is, tenzij het waterlichaam een uitgesproken lineair traject heeft (Packet et al., 2018, pp. 11-13). Niet afgedekte opvangbekkens bij serres, tuin- en landbouwbedrijven en overstromingsbekkens worden als waterbekken beschouwd, net als overstromingsbekkens en wadi's die langer waterhoudend zijn. Zwembaden en waterreservoirs die een onderdeel vormen van een industriele installatie (bv. waterzuivering) worden niet tot ecosysteem 9 gerekend. Als het meer of waterbekken zich bevindt op een perceel dat tot ecosysteem 1 behoort (bv. een tuinvijver of een park) of als het een recreatieve functie heeft, wordt het tot ecosysteem 1 gerekend.
**Bron:** Watervlakkenkaart INBO.}",
    r"{**10. Estuaria en overgangswateren**
**Beschrijving:** Getijdengevoelig deel van de IJzer en van de Schelde en haar zijrivieren (Vanlierde et al., 2014). De getijdengevoelige zone van het Schelde-estuarium omvat de volledige brakwater-zoetwatergradient (Jochems et al., 2002; Van den Bergh et al., 2009). In de IJzer beperkt de getijdengevoelige zone zich tot aan het sluizencomplex 'Ganzepoot'. De estuaria omvatten naast de loop van de rivier ook de slikken en schorren, tot en met de supralitorale zone die op basis van tijdata gemiddeld vier maal per jaar overspoelt (Van Ryckegem et al., 2025). Ook het water binnen de kaaimuren van de zeehavens (Nieuwpoort, Blankenberge, Oostende) wordt tot de overgangswateren gerekend. Het Zwin beantwoordt aan de EU-definitie van ecosysteemtype '10.2 Estuaries and bays' en wordt daarom tot ecosysteem 10 gerekend. Het laagstrand, het gebied tussen de gemiddelde hoogwaterlijn en de gemiddelde laagwaterlijn (Vriens et al., 2011), beantwoordt aan de EU-definitie van ecosysteemtype '10.3.1 Intertidal flats' en wordt eveneens tot ecosysteem 10 gerekend. Dit geldt ook voor de slikken in het Zwin. In tegenstelling tot de brak- en zoetwaterschorren langsheen rivieren onder getijdeninvloed, staan de mariene schorren nabij de baai van Heist en in het Zwin rechtstreeks onder invloed van zeewater. Zij beantwoorden aan de EU-definitie van 11.4.1 Coastal saltmarshes en worden tot ecosysteem 11 gerekend.
**Bron:** Het aquatisch deel van het getijdengevoelige deel van de Schelde en zijrivieren wordt bepaald op basis van de datalaag Watergang (Wtz) uit het GRB (Poelmans et al., 2022, pp. 40-41) en op basis van MONEOS-data, fysiotopen 'ondiep subtidaal', 'matig diep subtidaal', 'diep subtidaal' en 'zeer diep subtidaal' (Poelmans et al., 2022, pp. 40, 46). Het terrestrisch deel van het Schelde-estuarium is bepaald op basis van MONEOS-data, fysiotopen hoogslik, middelhoog slik en laag slik (voor de slikken) en hoog supralitoraal, supralitoraal (voor de schorren). De slikken en schorren van het IJzerestuarium zijn gebaseerd op de BWK, vegetatietypes ds en da (enkel buitendijks). De slikken van het Zwin zijn bepaald op basis van de BWK, vegetatietype ds. Het laagstrand is bepaald op basis van de BWK, vegetatietypes dl (strand, indien het veld INFO 'laagstrand' aangeeft), en dz (zandbank).}",
    r"{**11. Stranden, duinen en wetlands aan de kust**
**Beschrijving:** Het strand vanaf de gemiddelde hoogwaterlijn (hoogstrand) en de kustduinen, met inbegrip van stuifduinen, duinen begroeid met lage vegetatie en struiken, en vochtige duinpannen, in zoverre zij niet op percelen van het ruimtebeslag (ecosysteem 1) liggen of op percelen met een professionele landbouwfunctie. Beboste duinen worden tot ecosysteem 4 gerekend.
**Bron:** Landgebruiksbestand Vlaanderen, niveau 1 (Kustduin) en Biologische Waarderingskaart. Voor een specifieke lijst van strand- en kustduinvegetatie, zie bijlage.}",
    r"{**12. Mariene ecosystemen (kustwateren, continentaal plat en open oceaan)**
**Beschrijving:** De zee vanaf de gemiddelde laagwaterlijn, met inbegrip van de Belgische territoriale wateren (12 zeemijl vanaf de kust) en de exclusieve economische zone (EEZ). De gemiddelde laagwaterlijn wordt afgeleid van de BWK en verandert continu door een wisselwerking van stranderosie en strandsuppletie.
**Bron:** <ref toevoegen>}"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

kable(ecosystem_types, format = "html", align = "l", escape = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  ) |>
  column_spec(1, width = "18%") |>
  column_spec(2, width = "41%") |>
  column_spec(3, width = "41%")
```

```{r}
#| label: tbl-settlement
#| tbl-cap: EU and Flemish level-2 ecosystem type definitions for class 1 (Settlements and other artificial areas). Flemish columns are in Dutch, pending review and translation. Level-2 class 1.6 has no EU-level counterpart.

ecosystem_types_lvl2 <- data.frame(
  `EU Ecosystem Type level 2` = c(
    "1.1 Continuous settlement area",
    "1.2 Discontinuous settlement area",
    "1.3 Infrastructure and industrial areas",
    "1.4 Urban greenspace",
    "1.5 Other artificial areas",
    ""
  ),
  `EU Description and Correspondence` = c(
    r"{**Description:** Also designated as continuous urban area or fabric. The continuous settlement area type is assigned when settlement structures and transport networks are dominating the surface area. At least 80% of the land surface in the ecosystem asset is covered by impermeable features such as buildings, roads and artificially surfaced areas. **Source:** CORINE 1.1.1. Continuous urban fabric}",
    r"{**Description:** Also designated as discontinuous urban area or fabric. The discontinuous settlement area type is assigned when settlement structures and transport networks associated with vegetated areas and bare surfaces are present and occupy significant surfaces in a discontinuous spatial pattern. The impermeable features such as buildings, roads and artificially surfaced areas range from 30 to 80% land coverage in the ecosystem asset. **Correspondence:** CORINE 1.1.2. Discontinuous urban fabric}",
    r"{**Description:** This includes land used for transport infrastructure such as roads, airports and ports, including associated areas (e.g. planted trees and bushes lines or spontaneous vegetation, such as grass verges), and artificial areas other than buildings that are used for commercial and industrial purposes including extraction sites, dump areas, and construction sites. **Correspondence:** CORINE 1.2. Industrial, commercial and transport, 1.3 Mine, dump and construction sites}",
    r"{**Description:** Areas with vegetation within or partly embraced by urban fabric. This class is assigned for urban greenery, which usually has recreational or ornamental character and is usually accessible for the public. Includes small urban water bodies. Areas with soil sealing above 30% should be moved to class 1.1 or 1.2. **Correspondence:** CORINE Primarily 1.4.1 Green urban areas, (1.4.2 Sport and leisure facilities)}",
    r"{**Description:** Areas that are artificial but do not fall under the other level 2 categories of settlements and artificial areas. This includes e.g. permanent green houses, cemeteries even if predominantly green. Countries should also record here types of artificial land of specific national importance, such as archaeological sites or others, if they do not belong to any of the level 2 categories 1.1-1.4. Countries should inform what they cover in this category in a footnote in the questionnaire. **Correspondence:** CORINE Components of: 1.4.2 Sport and leisure facilities, 1.4.1 Green urban areas, 1.1.2 Discontinuous urban fabric}",
    ""
  ),
  `Flemish Ecosystem Type level 2` = c(
    "1.1 Gebouwen in woongebieden en andere kunstmatige gebieden",
    "1.2 Andere verharding in woongebieden en andere kunstmatige gebieden",
    "1.3 Laag groen in woongebieden en andere kunstmatige gebieden",
    "1.4 Hoog groen in woongebieden en andere kunstmatige gebieden",
    "1.5 Stilstaand water in woongebieden en andere kunstmatige gebieden",
    "1.6 Onduidelijke bodembedekking in woongebieden en andere kunstmatige gebieden"
  ),
  `Flemish Description and Correspondence (in Dutch, to be reviewed, translation after review)` = c(
    r"{**Description:** Gebouwen met een residentiele, commerciele, industriele, openbare, recreatieve of transportfunctie, voor een andere vorm dienstverlening of voor landbouwdoeleinden, met inbegrip van permanente serres en landbouwinfrastructuur. Gebouwen omvatten ook wat het Grootschalig Referentiebestand (GRB) beschrijft als 'kunstwerken (KNW)' (Poelmans et al., 2022).
**Correspondence:** Landgebruiksbestand Vlaanderen, niveau 1 (Gebouw).}",
    r"{**Description:** Door de mens gecreeerde verharding, andere dan gebouwen, op bebouwde percelen of als onderdeel van transportinfrastructuur.
**Correspondence:** Landgebruiksbestand Vlaanderen, niveau 1 (Weg, Spoorweg en een deel van Overig), ofwel > 75% verhard volgens de Jaarlijkse Bodemafdekkingskaart.}",
    r"{**Description:** Alle delen van percelen met ruimtebeslag die minder dan 76% verhard zijn en met vegetatie lager dan 3 meter of een niet-begroeide bodem.
**Correspondence:** Landgebruiksbestand Vlaanderen (Ruimtebeslag) en Groenkaart (vegetatie < 3m).}",
    r"{**Description:** Alle delen van percelen met ruimtebeslag die minder dan 76% verhard zijn en waarop bos of vegetatie hoger dan 3 meter voorkomt.
**Correspondence:** Landgebruiksbestand Vlaanderen (Ruimtebeslag) en Groenkaart (vegetatie > 3m).}",
    r"{**Description:** Alle zones van percelen met ruimtebeslag met duidelijk begrensd stilstaand open water dat gedurende meerdere jaren een groot deel van het vegetatieseizoen aanwezig is. Het waterlichaam wordt als 'stilstaand' beschouwd als er geen visueel waarneembare stroming is, tenzij het waterlichaam een uitgesproken lineair traject heeft (Packet et al., 2018, pp. 11-13). Niet afgedekte opvangbekkens bij serres, tuin- en landbouwbedrijven en overstromingsbekkens worden als waterbekken beschouwd, net als overstromingsbekkens en wadi's die langer waterhoudend zijn. Zwembaden en waterreservoirs die een onderdeel vormen van een industriele installatie (bv. waterzuivering) worden niet tot ecosysteem 1.5 gerekend maar tot 1.2 of 1.1.
**Correspondence:** Landgebruiksbestand Vlaanderen (Ruimtebeslag) en Watervlakkenkaart.}",
    r"{**Description:** Deel van het ruimtebeslag dat niet met zekerheid tot ecosysteem 1.1 tot 1.5 kan worden gerekend. Doorgaans gaat het om zones waarop zich verdorde vegetatie bevindt die niet als 'groen' wordt herkend op de groenkaart, en/of met een gedeeltelijke verharding < 76%, en waarvoor ook de BWK geen gegevens bevat.
**Correspondence:** Landgebruiksbestand Vlaanderen (Ruimtebeslag), Jaarlijkse Bodemafdekkingskaart (< 76% verharding), Groenkaart (geen laag of hoog groen) en BWK (geen gegevens).}"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

kable(ecosystem_types_lvl2, format = "html", align = "l", escape = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  ) |>
  column_spec(1, width = "14%") |>
  column_spec(2, width = "36%") |>
  column_spec(3, width = "14%") |>
  column_spec(4, width = "36%")
```

### Choice of Sampling Unit

For each selected raster cell (10 m $\times$ 10 m) (GRTS sample selection procedure), we defined: 

- **Primary Sampling Unit (PSU):** A square polygon (block) with dimensions of $5 \times 5$ raster cells (centered around the GRTS selected raster cell).
- **Secondary Sampling Units (SSU):** All 10 m $\times$ 10 m raster cells within the PSU.

In practice, the entire PSU block is digitized.
Digitization consists of subdividing the 50 m $\times$ 50 m polygon (PSU) into smaller polygons, with a primary and an (optional) alternative label assigned to each polygon).
Each subdivision is determined by desktop interpretation of orthophotos corresponding to the years for which FLEA maps are available (2016 - 2019 - 2021).
Every change of ecosystem type within a given year is a subdivision and every change of ecosystem type between years is a subdivision (see @fig-example-polygon).
The resulting attribute table for each subdivision polygon needs to be filled in with validation ecosystem labels for 2016, 2019 and 2021.

To speed up the validation work, the 50 m $\times$ 50 m polygons are subdivided and pre-labelled programmatically based on source layers for ecosystem types that can be easily recognized, such as buildings.
The programmatic pre-labelling must be validated as well.

This approach gives us a lot of freedom in the analysis stage (see @sec-analysis).

```{r}
#| label: fig-example-polygon
#| fig-cap: Example of a 50 m x 50 m validation polygon and how it will be subdivided based on changes in ecosystem types between and within years.

psu_matrix <- matrix(
  c(0, 0, 50, 0, 50, 50, 0, 50, 0, 0), ncol = 2, byrow = TRUE
)
psu <- st_polygon(list(psu_matrix)) %>% st_sfc()

rast10 <- sf::st_make_grid(psu, offset = c(0, 0), n = 5)

# Create mock ecosystem data for 2016
p16_1 <- st_polygon(list(matrix(c(0, 0, 25, 0, 25, 50, 0, 50, 0, 0),
                                ncol = 2, byrow = TRUE)))
p16_2 <- st_polygon(list(matrix(c(25, 0, 50, 0, 50, 50, 25, 50, 25, 0),
                                ncol = 2, byrow = TRUE)))
eco_2016 <- st_sf(
  ecosystem = c("Forest", "Grassland") |>
    factor(levels = c("Built-up", "Forest", "Agriculture", "Grassland")),
  jaar = 2016,
  geometry = st_sfc(p16_1, p16_2)
)

# ecosystem data for 2019
p19_1 <- st_polygon(list(matrix(c(0, 0, 50, 0, 50, 30, 0, 30, 0, 0),
                                ncol = 2, byrow = TRUE)))
p19_2 <- st_polygon(list(matrix(c(0, 30, 50, 30, 50, 50, 0, 50, 0, 30),
                                ncol = 2, byrow = TRUE)))
eco_2019 <- st_sf(
  ecosystem = c("Forest", "Agriculture") |>
    factor(levels = c("Built-up", "Forest", "Agriculture", "Grassland")),
  jaar = 2019,
  geometry = st_sfc(p19_1, p19_2)
)

# ecosystem data for 2021
# Scenario: A 15x15m building is constructed in the bottom-left corner.
# The rest of the PSU is mapped as Forest.
p21_bldg <- st_polygon(
  list(matrix(c(0, 0, 15, 0, 15, 15, 0, 15, 0, 0),
              ncol = 2, byrow = TRUE))
  ) %>% st_sfc()
p21_rest <- st_difference(eco_2019$geometry, p21_bldg) # The rest of the PSU
eco_2021 <- st_sf(
  ecosystem = c("Built-up", "Forest", "Agriculture") |>
    factor(levels = c("Built-up", "Forest", "Agriculture", "Grassland")),
  jaar = 2021,
  geometry = c(p21_bldg, p21_rest)
)

# Calculate the intersections (The programmatic subdivision)
inter_16_19 <- st_intersection(eco_2016, eco_2019) |> suppressWarnings()
final_subdivisions <- st_intersection(inter_16_19, eco_2021) |>
  suppressWarnings() |>
  st_collection_extract("POLYGON")
final_subdivisions$poly_id <- seq_len(nrow(final_subdivisions))
final_subdivisions$area_m2 <- st_area(final_subdivisions)
final_subdivisions <- final_subdivisions |>
  mutate(
    conversion = sprintf(
      "%s%s%s",
      str_sub(ecosystem, end = 1),
      str_sub(ecosystem.1, end = 1),
      str_sub(ecosystem.2, end = 1)
      )
  )

colours <- c(
  "Forest" = topo.colors(4)[1],
  "Grassland" = topo.colors(4)[2],
  "Agriculture" =  topo.colors(4)[3],
  "Built-up" = topo.colors(4)[4]
)
my_levels <- c("Built-up", "Forest", "Agriculture", "Grassland")
ordered_colours <- unname(colours[my_levels]) 

dummy_df <- data.frame(
  ecosystem = factor(
    c("Built-up", "Forest", "Agriculture", "Grassland"), 
    levels = c("Built-up", "Forest", "Agriculture", "Grassland")
  )
)

shared_scale <- scale_fill_manual(
  name = "Ecosystem",
  values = colours,
  drop = FALSE
)

# Plotting the results using ggplot2 and patchwork
# Map the actual data, but TURN OFF its legend
plot_2016 <- ggplot() +
  geom_sf(
    data = eco_2016,
    aes(fill = ecosystem),
    alpha = 0.7,
    show.legend = TRUE) +
  geom_sf(data = rast10, fill = NA, linetype = "dotted") +
  shared_scale +
  theme_void() +
  ggtitle("2016 ecosystems")

plot_2019 <- ggplot() +
  geom_sf(data = eco_2019,
          aes(fill = ecosystem),
          alpha = 0.7,
          show.legend = TRUE) +
  geom_sf(data = rast10, fill = NA, linetype = "dotted") +
  shared_scale +
  theme_void() +
  ggtitle("2019 ecosystems")

plot_2021 <- ggplot() +
  geom_sf(data = eco_2021,
          aes(fill = ecosystem),
          alpha = 0.7,
          show.legend = TRUE) +
  geom_sf(data = rast10, fill = NA, linetype = "dotted") +
  shared_scale +
  theme_void() +
  ggtitle("2021 ecosystems")

plot_final <- ggplot(final_subdivisions) +
  geom_sf(fill = "gray90", color = "black", linewidth = 0.8) +
  geom_sf_text(aes(label = conversion), size = 4, fontface = "bold") +
  geom_sf(data = rast10, fill = NA, linetype = "dotted") +
  theme_void() +
  ggtitle("Final subdivisions", subtitle = "Intersection of all years")

# Combine into a 2x2 grid
plot_2016 + plot_2019 + plot_2021 + plot_final +
  plot_layout(ncol = 2, guides = "collect") 
```


### Reference data used

Regarding the **resolution of the aerial photographs (orthophotos)**, the medium-scale photographs from 2023 have a resolution of 15 cm.
In the year 2000, this was still 1 m.
Therefore, it is possible that the resolution evolved during the period from 2013 to 2022.
As a result, photographs used to validate a 2013 ecosystem map might be less accurate than those available for 2022.

## Analysis

Due to the more complex sampling and response design (balancing precision and cost-efficiency), the standard formulas provided in @olofsson2014 will only apply at the PSU level.
However, this sampling design retains the flexibility to calculate accuracy and area estimates for different scenarios in post-processing.

For instance, if we define a PSU as a block of 5 $\times$ 5 raster cells, we can determine the label at the PSU level for both the reference data and the map, and subsequently process them using the formulas from @olofsson2014.

We can also determine the label on a subset of the PSU, such as the central 3 $\times$ 3 block of raster cells, and process it identically.

We can also utilize all information (from all raster cells) to perform validation at the raster cell level; in this case, formulas from model-assisted survey sampling literature and/or estimators for cluster sampling will likely be required for the analysis.
@cochran1977 provides the formula to determine the effect of cluster size when using cluster sampling to estimate a proportion:

$$\begin{aligned}
\text{total sum of squares} &= \text{within sum of squares} + \text{between sum of squares} \\
NMP(1-P) &= M\sum_{i=1}^n{(p_i - P)^2} + Mp_i(1-p_i)
\end{aligned}$$

This relationship can be used to calculate the various variance components as a function of the cluster size $M$.
This allows us to evaluate whether an optimal cluster size exists purely in terms of variance (disregarding costs).

In all cases, we will need rules for determining the reference labels.
Since our response design results in a vector-based ground truth, we need to translate the vector polygon data (from the digitized mapping exercise) to a raster cell or an aggregate of raster cells (3 x 3, 5 x 5).
This will be done by assigning the label from the polygon with the largest intersecting area with the (aggregate of) raster cell(s).
In case of polygons with equal intersecting area, a label will be selected randomly from these polygon labels.

